TMR: Detecting Structure in Graphical Data

### 2021-01-20 interim-report/interim-report.tex
Final draft

%custom commands
\newcommand{\dft}{\mathcal{F}}
\newcommand{\idft}{\mathcal{F}^{-1}}
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\cmplx}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\Gcal}{\mathcal{G}}
\newcommand{\Vcal}{\mathcal{V}}
\newcommand{\Ecal}{\mathcal{E}}
\newcommand{\pbold}{\boldsymbol{p}}
\newcommand{\sigmabold}{\boldsymbol{\sigma}}
\newcommand{\lik}{\mathcal{L}}
\newcommand{\kl}{\mathcal{D}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\SBM}{\textrm{SBM}}
\newcommand{\iid}{\stackrel{iid}{\sim}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\prob}{\mathbbm{P}}
\newcommand{\figwidth}{0.7\linewidth}

@@

\begin{abstract}
We produce vast quantities of graphical data every day and yet the techniques used to analyse their structure are still in their infancy. We begin with the definition of the Stochastic Block Model (SBM) widely used in academia and develop a theorem to verify structure in a labelled graph through a hypothesis test. We explore the results of applying this theorem to Facebook friendship egonets and find that gender does indeed influence how friendships form on Facebook. Nevertheless, this binary yes-no result is unsatisfactory.

To rank order vertex labels by importance we instead approach the problem from the opposite direction: given a graph we detect the partitions that are in some sense typical and then develop a classifier to map from vertex features to partition. The classifier then implicitly rank orders the input features automatically. Nevertheless, work must still be done to refine this classifier and tie it back to the rigorous hypothesis testing framework already developed.
\end{abstract}

@@

Furthermore, we spot the following property $f(w, e_1, m_1) \cdot f(w, e_2, m_2) = f(w, e_1 + e_2, m_1 + m_2)$ or in other words, the function $f$ is linear in its second and third arguments given the same first argument. As such we can manipulate equation \ref{eqn:test-statistic-start} greatly to give:

@@

Where $\hat{p} \coloneqq E_\alpha / M_\alpha$, $\hat{q} \coloneqq E_\beta / M_\beta$ and $\hat{r} \coloneqq (E_\alpha + E_\beta) / (M_\alpha + M_\beta)$. These are essentially maximum-likelihood estimates of parameters of $W$ constrained by the corresponding hypothesis. Note that the max-likelihood estimate of $W_\alpha$ under $H_1$ is denoted by the symbol $\hat{p}$ and should not be confused with the community prior $p$ though context will make the distinction apparent. We now state lemma \ref{lem:kl-div}.

@@

	where $\kl(g || h)$ is the Kullback-Leibler divergence between two probability mass functions $g, h: \Xcal \mapsto [0, 1]$. $\kl$ is defined in discrete space as $\kl(g || h) \coloneqq \sum_{x \in \Xcal} g(x) \log \frac{g(x)}{h(x)}$ and $\textrm{Bern(p)}$ denotes the Bernoulli p.m.f with parameter $p$.
	\label{lem:kl-div}
\end{lemma}

Proving lemma \ref{lem:kl-div} is simply a case of algebraic manipulation and is omitted here. However, this allows us to simplify the test-statistic into a form that is easier to compute (equation \ref{eqn:test-stat-compute-form}). However, we must simplify further still using lemma \ref{lem:kl-approx-chi}.
%
\begin{equation}
	\begin{aligned}
	t_n = &M_\alpha \kl\left( \textrm{Bern}(\hat{p}) || \textrm{Bern}(\hat{r})\right) \\
	&+ M_\beta \kl\left( \textrm{Bern}(\hat{q}) || \textrm{Bern}(\hat{r})\right)
	\end{aligned}
	\label{eqn:test-stat-compute-form}
\end{equation}
%
\begin{lemma}[$\chi^2$ approximation]
	\label{lem:kl-approx-chi}
	For $g(x) \approx h(x)$ then $\kl(g || h) \approx \frac{1}{2} \chi^2(g || h)$ where $\chi^2$ is the chi-squared distance between two distributions defined as $\chi^2(g || h) \coloneqq \sum_{x \in \Xcal} \frac{(g(x)-h(x))^2}{h(x)}$ 
\end{lemma}

The proof requires defining $\delta(x) \coloneqq g(x) - h(x)$ and taking a Taylor expansion of $\log (1 + \delta / h)$ - omitted for bervity. The chi-squared distance between two Bernoullis $g = Bern(p)$ and $h = Bern(q)$ has a very simple form:
%
\begin{equation}
\chi^2(\textrm{Bern}(p) || \textrm{Bern}(q)) = \left( \frac{p-q}{\sqrt{q(1-q)}} \right)^2
\end{equation}
%
We can apply these results to equation \ref{eqn:test-stat-compute-form} to obtain an approximate expression for $t_n$ under the null. Under $H_0$ and for large $n$, the estimates $\hat{p}$, $\hat{q}$ and $\hat{r}$ are all close meaning Lemma \ref{lem:kl-approx-chi} is applicable. This gives:
%
\begin{align}
	t_n &\approx \frac{1}{2 \hat{r}(1 - \hat{r})} \left( 
	M_\alpha \left(\hat{p} - \hat{r}\right)^2 +
	M_\beta \left(\hat{q} - \hat{r} \right)^2 
	\right) \nonumber \\
	&= \frac{1}{2} \left( \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1- \hat{r})(1/M_\alpha + 1/M_\beta)}} \right)^2
	= \frac{1}{2} z_n^2
\end{align}
%
Through application of the Central Limit Theorem (CLT) and Slutsky's Theorem it can be shown that as $n \rightarrow \infty$ under the null, the term in brackets on the final line is distributed like a standard Gaussian $z_n \sim \Gaussian(0,1)$. The proof is long-winded and therefore omitted from this short report. The result leads to theorem \ref{theorem:hyp-test-sbm-chi}.

\begin{theorem}[$\chi^2$ test]
	For $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$, given the realised graph and class labels $(X, \Gcal)$ we can perform a hypothesis test on entries $W_\alpha$ and $W_\beta$ of the connectivity matrix $W$.
	%
	\begin{align*}
	H_0:& \quad W_{\alpha} = W_{\beta} \\
	H_1:& \quad W_{\alpha} \neq W_{\beta}
	\end{align*}
	If the log-likelihood ratio test statistic $t_n$ is computed as in equation \ref{eqn:test-stat-compute-form}, then as the number of vertices $n \rightarrow \infty, t_n \sim \frac{1}{2} \chi^2_1$ under the null $H_0$. Therefore, we reject $H_0$ at the $100(\zeta)\%$ confidence level if and only if $2t_n \geq \psi^{-1}(\zeta)$, where $\psi^{-1}$ is the $\chi^2_1$ inverse cdf satisfying $\prob(Y \leq \psi^{-1}(\zeta)) = \zeta$ given $Y \sim \chi^2_1$.

@@

We seek to apply theorem \ref{theorem:hyp-test-sbm-chi} to real-world graphical datasets. We start by analysing social network graphs. The Stanford Network Analysis Project (SNAP) \cite{snapnets} offers a wealth of Facebook egonets. An egonet is simply a graph where all vertices (in this case Facebook users) are connected to one central node (the ego-node). The data consists of the undirected set of edges $\Gcal$ indicating whether any two vertices are connected (friends on Facebook) and matrix $F$ containing binary feature flags for each vertex. Example feature flags are given below: \\

@@

We use an SBM model with $k=2$ communities (1: gender-77 on and 2: gender-77 off) to model the egonet and perform a three-way hypothesis test on the parameters of the connectivity Matrix $W$.
%
\begin{equation}
	W = \begin{bmatrix}
		p_1 & q \\
		q & p_2
	\end{bmatrix}
\end{equation}
%
Therefore, $p_1$ is the probability that two vertices of gender-77 are connected, $q$ is the crossover probability and $p_2$ is the connection probability within the gender-78 community
\footnote{Strictly speaking this is the not gender-77 community as there are a handful of vertices that do not disclose their gender to Facebook or do not fall into this binary}
. The results of the hypothesis tests are given in table \ref{tab:egonet-0-hyp-tests} alongside p-values (we choose a 95\% significance level to reject the null). However, in some cases the test statistics were so extreme that the p-value saturated to 0.
%
\begin{table}[H]
	\centering
	\begin{tabular}{c c c c c}
		$n_1$                & $n_2$                & $\hat{p_1}$ & $\hat{q}$ & $\hat{p_2}$       \\ \hline
		130 & 218 & $8.13 \times 10^{-2}$ & $7.66 \times 10^{-2}$ & $10.6 \times 10^{-2}$
	\end{tabular}
	\caption{Egonet-0 properties and parameter estimates}
	\label{tab:egonet-0-props}
\end{table}
%
\begin{table}[H]
	\centering
	\begin{tabular}{cccccc}
		Test & $H_0$                & $H_1$                & p-value & $H_0$ rejected       \\ \hline
		1    & $p_1 = q$            & $p_1 \neq q$         & 0.158 &  No                \\
		2    & $p_1 = p_2$ & $p_1 \neq p_2$ & 0.000 & Yes \\
		3    & $q = p_2$ & $q \neq p_2$ &  0.000 & Yes
	\end{tabular}
	\caption{Egonet-0 hypothesis test results}
	\label{tab:egonet-0-hyp-tests}
\end{table}

@@

However, the exercise has highlighted some interesting points. The hypothesis test framework does not quantify the magnitude of the difference just whether or not a difference exists. Furthermore, The method relies on the analyst to specify the features of interest so results may be misleading if third variables are missing from the analysis.

To address these shortcomings we instead approach the problem from a different angle. For now we have been given a graph with fully labelled vertices and ask if a given feature affects the structure. Instead, we could take a graph and partition first (without using feature information) and then ask which features explain the partition. For that we need a way of detecting structure.

@@

Community detection is the problem of recovering the communities from the graph $\Gcal$. There are 4 recovery regimes (plus the trivial no recovery), as defined by Abbe \cite{Abbe}. We choose to focus on weak recovery as it is the least strict of all. Therefore, there are very few constraints on the SBM parameters and the theory can be applied to a wide range of graphs. A statement of weak recovery is given in definition \ref{defn:weak-recovery} (adapated from \cite{Abbe}).

@@

	Where $o(1)$ denotes a family of functions that tend to $0$ as $n$ gets large. We recall the definition of $\Omega_i = \Omega_i(X) \coloneqq \{v \in \Vcal : X_v = i\}$. In other words, we require the partition $S \subset \Vcal$ to cut across communities such that the fraction of a particular community in partition $S$ with respect to the whole community is different for a choice of two communities.  
\end{definition}

One of the most promising weak recovery algorithms is called Acyclic Belief Propagataion (ABP). We implement a linearised version adapted from Abbe and Sandon \cite{Linear-ABP}. This is a message-passing algorithm that solves weak recovery in the sense of definition \ref{defn:weak-recovery}. The original paper goes into much more detail than we have space to discuss here. Nevertheless, the premise is that one can infer the community of a particular vertex simply from the labels of all its neighbours. The algorithm is non-deterministic and requires the graph $\Gcal$ as well as two hyper-parameters $r$ (the maximum cyclic length to correct for) and $T$ (the number of iterations).

@@

We colour each vertex according to this quantity $\tilde{\sigma}_v$ and plot the graph
\footnote{ABP is non-deterministic so each figure is the result of a different run of ABP. No exact comparison is possible between figures but the results are in some sense typical.}
on figure \ref{fig:abp-output-norm}. We see intuitively that ABP has worked. The blue cluster in the lower right is clearly part of the same community. Furthermore, many vertices that do not seem to belong to any cluster have uncertain classifications (the colour is close to green). Inspecting figure \ref{fig:ego-0-by-gender} shows us that the sets $S$ and $S^C$ do separate nicely with high intra-group connectivity and lower inter-group connection density. After all, we seek to find a roughly balanced partition that minimises the number of edges between the two sets.
%
\begin{figure}[H]
	\centering
	\includegraphics[width=\figwidth]{ego-0-abp-by-gender.png}
	\caption{Vertices grouped by ABP partition: set $S$ (left cluster) and set $S^C$ (right cluster)}
	\label{fig:abp-output-by-gender}
\end{figure}

Within set $S$ we have $\{58, 122\}$ 77-ers and 78-ers respectively and for $S^C$ the breakdown is $\{72, 96\}$. Therefore, the fraction of each gender cohort in set S can be calculated:
%
\begin{equation}
	\begin{aligned}
	\frac{|\Omega_{77} \cap S|}{|\Omega_{77}|} &= \frac{58}{58+72} = 0.446 \\
	\frac{|\Omega_{78} \cap S|}{|\Omega_{78}|} &= \frac{122}{122+96} = 0.560
	\end{aligned}
\end{equation}
%
Clearly the fractions of each community in the partition $S$ differ substantially so we say that weak recovery is solved. The set $S$ contains a higher proportion of the overall 78-population than of the 77-population.

@@

This part of the project is still in its early stages so these are just preliminary results. We necessarily remove the constraint on the community sets $\Omega_i$ to be disjoint as one vertex can have multiple features turned on. We perform a linear regression on the $n \times 1$ normalised belief vector $\tilde{\sigma}$ with the $n \times f$ feature matrix $F$ as the explanatory variable. $F_{ij} = 1$ if vertex $i$ has feature $j$ turned on and $0$ if feature $j$ is off. The equation we try to fit is given below:
%
\begin{equation}
	\tilde{\sigma} = a + F b + \epsilon
	\label{eqn:lin-regression}
\end{equation}
%
Linear regression finds the constant bias $a$ and $f \times 1$ weight vector $b$ that minimises the mean squared prediction error $\epsilon^T \epsilon / n$. The theory for linear regression is well known so omitted here (see \cite{yale-linear-regression} for a reference). We restrict $F$ to only include gender and language features and plot the regression coefficient for each feature in order of decreasing magnitude (figure \ref{fig:abp-output-lin-regression}).
%
\begin{figure}[H]
	\includegraphics[width=\linewidth]{ego-0-abp-lin-regression.png}
	\caption{Egonet-0 linear regression of $\tilde{\sigma}_v$ on select features ($R^2 = 0.113$)}
	\label{fig:abp-output-lin-regression}
\end{figure}

@@

This analysis is of course rather crude as we have failed to take into account multi-collinearity or even determined if a linear regression model is appropriate; indeed, an $R^2$ value of just 0.113 would suggest that it is not. Nevertheless, this serves as the platform for subsequent more rigorous analysis. Furthermore, we have not yet proven the conjecture that the regression coefficients are valid measures of causal impact on graphical structure.

\section{Future Direction}

The immediate next steps for the project are to refine the classification approach outlined in section \ref{sect:detection-to-inference}. Firstly, it should be tied back to the hypothesis testing framework to make sure that the results are proven to be meaningful. Secondly, the exact form of the classifier can be explored; a linear regression model is a good starting point but we can expand to a softmax neural network as that would allow us to generalise to more than just two partitions.

Indeed, ABP is severely limited by the fact that it only outputs 2 sets. Though weak recovery is solved, the partition is often hard to interpret as various clusters get subsumed into one. We instead would prefer to investigate algorithms that can identify arbitrarily many clusters. Force-directed approaches are the standard in industry for their speed of execution. One of the most common is the Fruchterman-Reingold algorithm which treats vertices as repelling objects and edges as restoring forces. It can then perform iterative updates on vertex positions until convergence or termination. This algorithm is what is used by the NetworkX package \cite{networkx} to compute graph layouts as seen in figures \ref{fig:ego-0-by-gender} and \ref{fig:abp-output-norm}. Song and Bressan \cite{force-directed} have had great success using the Fruchterman-Reingold algorithm followed by k-means clustering for community detection.

Lastly, the analysis must of course be expanded to more datasets to benchmark efficacy against existing methods and ensure the approach is generally applicable. Work has already begun analysing the influence of gender on academic collaboration graphs provided by Aminer \cite{aminer}. Nevertheless, this dataset does not contain gender explicitly, only first-name. However, this is cross-referenced with a list of popular baby-names to get a good replacement (as suggested by West et al. \cite{gender-scholarship}). Furthermore, the seminal dataset of political blogs from the 2004 US election \cite{polblogs} which started much of the analysis of the SBM will of course be analysed.

@@

This concludes the Technical Milestone Report for the project thus far. We start by fully fleshing out the theoretical foundation of the Stochastic Block Model and hypothesis tests on connectivity parameters. Nevertheless, this approach is too narrow in scope to be a useful tool on its own. We therefore set about the problem from the opposite direction; rather than verify structure we are already given, we detect structure in the graph without using labels and seek to identify which labels best recreate the detected partition. Early results are promising but this approach needs to be given the same rigorous treatment as the hypothesis testing framework. That will be the emphasis of future work.

>>>>

%custom commands
\newcommand{\dft}{\mathcal{F}}
\newcommand{\idft}{\mathcal{F}^{-1}}
\newcommand{\Xcal}{\mathcal{X}}
\newcommand{\Dcal}{\mathcal{D}}
\newcommand{\cmplx}{\mathbb{C}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Gaussian}{\mathcal{N}}
\newcommand{\Gcal}{\mathcal{G}}
\newcommand{\Vcal}{\mathcal{V}}
\newcommand{\Ecal}{\mathcal{E}}
\newcommand{\pbold}{\boldsymbol{p}}
\newcommand{\sigmabold}{\boldsymbol{\sigma}}
\newcommand{\lik}{\mathcal{L}}
\newcommand{\kl}{\mathcal{D}}
\newcommand{\Integers}{\mathbb{Z}}
\newcommand{\SBM}{\textrm{SBM}}
\newcommand{\iid}{\stackrel{iid}{\sim}}
\newcommand{\one}{\mathbbm{1}}
\newcommand{\prob}{\mathbbm{P}}
\newcommand{\figwidth}{0.8\linewidth}

@@

\begin{abstract}
We produce vast quantities of graphical data every day and yet the techniques used to analyse their structure are still in their infancy. We begin with the Stochastic Block Model (SBM) widely used in academia and derive a theorem to verify structure in a labelled graph through a hypothesis test. We apply this theorem to Facebook egonets and find that gender does indeed influence how friendships form on Facebook. Nevertheless, this binary yes-no result is unsatisfactory.

We instead approach the problem from the opposite direction: given a graph we detect the partitions that are in some sense typical and then develop a classifier to map from vertex features to partition. The classifier then implicitly rank orders the input features by importance. Nevertheless, work must still be done to refine this classifier and tie it back to the rigorous hypothesis testing framework already developed.
\end{abstract}

@@

Furthermore, we spot the following property $f(w, e_1, m_1) \cdot f(w, e_2, m_2) = f(w, e_1 + e_2, m_1 + m_2)$ or in other words, the function $f$ is linear in its second and third arguments given the same first argument. As such, we can manipulate equation \ref{eqn:test-statistic-start} greatly to give:

@@

Where $\hat{p} \coloneqq E_\alpha / M_\alpha$, $\hat{q} \coloneqq E_\beta / M_\beta$ and $\hat{r} \coloneqq (E_\alpha + E_\beta) / (M_\alpha + M_\beta)$. These are essentially max-likelihood estimates of the parameters of $W$ constrained by the corresponding hypothesis. Note that the max-likelihood estimate of $W_\alpha$ under $H_1$ is denoted by the symbol $\hat{p}$ and should not be confused with the community prior $p$ though context will make the distinction apparent. We now state lemma \ref{lem:kl-div}.

@@

	where $\kl(g || h)$ is the Kullback-Leibler divergence between two probability mass functions (pmf's) $g, h: \Xcal \mapsto [0, 1]$. $\kl$ is defined in discrete space as $\kl(g || h) \coloneqq \sum_{x \in \Xcal} g(x) \log \frac{g(x)}{h(x)}$ and $\textrm{Bern(p)}$ denotes the Bernoulli pmf with parameter $p$.
	\label{lem:kl-div}
\end{lemma}

Proving lemma \ref{lem:kl-div} is simply a case of algebraic manipulation and is omitted here. The lemma allows us to simplify the test-statistic into a form that is easier to compute (equation \ref{eqn:test-stat-compute-form}).
%
\begin{equation}
	\begin{aligned}
	t_n = &M_\alpha \kl\left( \textrm{Bern}(\hat{p}) || \textrm{Bern}(\hat{r})\right) \\
	&+ M_\beta \kl\left( \textrm{Bern}(\hat{q}) || \textrm{Bern}(\hat{r})\right)
	\end{aligned}
	\label{eqn:test-stat-compute-form}
\end{equation}
%
However, we must simplify further using lemma \ref{lem:kl-approx-chi}.
%
\begin{lemma}[$\chi^2$ approximation]
	\label{lem:kl-approx-chi}
	For $g(x) \approx h(x)$ then $\kl(g || h) \approx \frac{1}{2} \chi^2(g || h)$ where $\chi^2$ is the chi-squared distance between two distributions defined as $\chi^2(g || h) \coloneqq \sum_{x \in \Xcal} \frac{(g(x)-h(x))^2}{h(x)}$ 
\end{lemma}

The proof requires defining $\delta(x) \coloneqq g(x) - h(x)$ and taking a Taylor expansion of $\log (1 + \delta / h)$ - omitted for brevity. The chi-squared distance between two Bernoullis $g = Bern(p)$ and $h = Bern(q)$ has a very simple form:
%
\begin{equation}
\chi^2(\textrm{Bern}(p) || \textrm{Bern}(q)) = \left( \frac{p-q}{\sqrt{q(1-q)}} \right)^2
\end{equation}
%
We can apply these results to equation \ref{eqn:test-stat-compute-form} to obtain an approximate expression for $t_n$ valid for large $n$ under the null hypothesis $H_0$. Under these conditions, the estimates $\hat{p}$, $\hat{q}$ and $\hat{r}$ are all close meaning Lemma \ref{lem:kl-approx-chi} is applicable. This gives:
%
\begin{align}
	t_n &\approx \frac{1}{2 \hat{r}(1 - \hat{r})} \left( 
	M_\alpha \left(\hat{p} - \hat{r}\right)^2 +
	M_\beta \left(\hat{q} - \hat{r} \right)^2 
	\right) \nonumber \\
	&= \frac{1}{2} \left( \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1- \hat{r})(1/M_\alpha + 1/M_\beta)}} \right)^2
	= \frac{1}{2} z_n^2
\end{align}
%
Through application of the Central Limit Theorem (CLT) and Slutsky's Theorem it can be shown that as $n \rightarrow \infty$ under the null $H_0$, the term in brackets on the final line is distributed like a standard Gaussian $z_n \sim \Gaussian(0,1)$. The proof is long-winded and therefore omitted from this short report. The result leads to theorem \ref{theorem:hyp-test-sbm-chi}.

\begin{theorem}[$\chi^2$ test]
	For $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$, given the realised graph and class labels $(X, \Gcal)$ we can perform a hypothesis test on entries $W_\alpha$ and $W_\beta$ of the connectivity matrix $W$:
	%
	\begin{align*}
	H_0:& \quad W_{\alpha} = W_{\beta} \\
	H_1:& \quad W_{\alpha} \neq W_{\beta}
	\end{align*}
	If the log-likelihood ratio test statistic $t_n$ is computed as in equation \ref{eqn:test-stat-compute-form}, then as the number of vertices $n \rightarrow \infty, t_n \sim \frac{1}{2} \chi^2_1$ under the null $H_0$. Therefore, we reject $H_0$ at the $100(\zeta)\%$ confidence level if and only if $2t_n \geq \psi^{-1}(\zeta)$, where $\psi^{-1}$ is the $\chi^2_1$ inverse cdf satisfying $\prob(Y \leq \psi^{-1}(\zeta)) = \zeta$ given $Y \sim \chi^2_1$.

@@

We seek to apply theorem \ref{theorem:hyp-test-sbm-chi} to real-world graphical datasets. We start by analysing social network graphs. The Stanford Network Analysis Project (SNAP) \cite{snapnets} offers a wealth of Facebook egonets. An egonet is simply a graph where all vertices (in this case Facebook users) are connected to one central node (the egonode). The data consists of the undirected set of edges $\Gcal$, indicating whether any two vertices are connected (friends on Facebook) and matrix $F$, containing binary feature flags for each vertex. Example feature flags are given below: \\

@@

We use an SBM with $k=2$ communities (1: gender-77 on and 2: gender-77 off) to model the egonet and perform a three-way hypothesis test on the parameters of the connectivity Matrix $W$.
%
\begin{equation}
	W = \begin{bmatrix}
		p_1 & q \\
		q & p_2
	\end{bmatrix}
\end{equation}
%
Therefore, $p_1$ is the probability that two vertices of gender-77 are connected, $q$ is the crossover probability and $p_2$ is the connection probability within the gender-78 community
\footnote{Strictly speaking this is the not gender-77 community as there are a handful of vertices that do not disclose their gender to Facebook or do not fall into this binary}
. The results of the hypothesis tests are given in table \ref{tab:egonet-0-hyp-tests} alongside p-values (we choose a 95\% significance level to reject the null). However, in some cases the test statistics were so extreme that the p-value saturated to 0.
%
\begin{table}[H]
	\centering
	\begin{tabular}{c c c c c}
		$n_1$                & $n_2$                & $\hat{p_1}$ & $\hat{q}$ & $\hat{p_2}$       \\ \hline
		130 & 218 & $8.13 \times 10^{-2}$ & $7.66 \times 10^{-2}$ & $10.6 \times 10^{-2}$
	\end{tabular}
	\caption{Egonet-0 properties and parameter estimates}
	\label{tab:egonet-0-props}
\end{table}
%
\begin{table}[H]
	\centering
	\begin{tabular}{cccccc}
		Test & $H_0$                & $H_1$                & p-value & $H_0$ rejected       \\ \hline
		1    & $p_1 = q$            & $p_1 \neq q$         & 0.158 &  No                \\
		2    & $p_1 = p_2$ & $p_1 \neq p_2$ & 0.000 & Yes \\
		3    & $q = p_2$ & $q \neq p_2$ &  0.000 & Yes
	\end{tabular}
	\caption{Egonet-0 hypothesis test results}
	\label{tab:egonet-0-hyp-tests}
\end{table}

@@

That said, the exercise has highlighted some interesting points. The hypothesis test framework does not quantify the magnitude of the difference just whether or not a difference exists. Furthermore, The method relies on the analyst to specify the features of interest so results may be misleading if third variables are missing from the analysis.

To address these shortcomings we instead approach the problem from a different angle. For now we have been given a graph with fully labelled vertices and ask if a given label affects the structure. Instead, we could take a graph and partition first (without using feature information) and then ask which features explain the partition. For that we need a way of detecting structure.

@@

Community detection is the problem of recovering the communities from the graph $\Gcal$. There are 4 recovery regimes, as defined by Abbe \cite{Abbe}. We choose to focus on weak recovery as it is the least strict of all. Therefore, there are very few constraints on the SBM parameters and the theory can be applied to a wide range of graphs. A statement of weak recovery is given in definition \ref{defn:weak-recovery} (adapted from \cite{Abbe}).

@@

	Where $o(1)$ denotes a family of functions that tend to $0$ as $n$ increases. We recall the definition of $\Omega_i = \Omega_i(X) \coloneqq \{v \in \Vcal : X_v = i\}$. In other words, we require the partition $S \subset \Vcal$ to cut across communities such that the fraction of a particular community in partition $S$ with respect to the whole community is different for a choice of two communities.  
\end{definition}

One of the most promising weak recovery algorithms is called Acyclic Belief Propagataion (ABP). We implement a linearised version adapted from Abbe and Sandon \cite{Linear-ABP}. This is a message-passing algorithm that solves weak recovery in the sense of definition \ref{defn:weak-recovery}. The original paper goes into much more detail than we have space to discuss here. Nevertheless, the premise is that one can infer the community of a particular vertex simply from the labels of all its neighbours. The algorithm is non-deterministic
\footnote{As ABP is non-deterministic, each figure is the result of a different run of ABP. No exact comparison is possible between figures but the results are in some sense typical.}
and requires the graph $\Gcal$ as well as two hyper-parameters $r$ (the maximum cyclic length to correct for) and $T$ (the number of iterations).

@@

We colour each vertex according to this quantity $\tilde{\sigma}_v$ and plot the graph on figure \ref{fig:abp-output-norm}. We see intuitively that ABP has worked. The blue cluster in the lower right is clearly part of the same community. Furthermore, many vertices that do not seem to belong to any cluster have uncertain classifications (the colour is close to green). Inspecting figure \ref{fig:ego-0-by-gender} shows us that the sets $S$ and $S^C$ do separate nicely with high intra-group connectivity and lower inter-group connection density. Indeed, an alternative interpretation of weak recovery is finding the partition that minimises the number of edges between the two sets.
%
\begin{figure}[H]
	\centering
	\includegraphics[width=\figwidth]{ego-0-abp-by-gender.png}
	\caption{Vertices grouped by ABP partition: set $S$ (left cluster) and set $S^C$ (right cluster)}
	\label{fig:abp-output-by-gender}
\end{figure}

Within set $S$ we have $\{58, 122\}$ 77-ers and 78-ers respectively and for $S^C$ the breakdown is $\{72, 96\}$. Therefore, the fraction of each gender cohort in set S can be calculated:
%
\begin{equation}
	\begin{aligned}
	\frac{|\Omega_{77} \cap S|}{|\Omega_{77}|} &= \frac{58}{58+72} = 0.446 \\
	\frac{|\Omega_{78} \cap S|}{|\Omega_{78}|} &= \frac{122}{122+96} = 0.560
	\end{aligned}
\end{equation}
%
Clearly the fractions of each community in the partition $S$ differ substantially so we say that weak recovery is solved in the sense of definition \ref{defn:weak-recovery}. The set $S$ contains a higher proportion of the overall 78-population than of the 77-population.

@@

This part of the project is still in its early stages so we present preliminary results. We necessarily remove the constraint on the community sets $\Omega_i$ to be disjoint as one vertex can have multiple features turned on. We perform a linear regression on the $n \times 1$ normalised belief vector $\tilde{\sigma}$ with the $n \times f$ feature matrix $F$ as the explanatory variable. $F_{ij} = 1$ if vertex $i$ has feature $j$ turned on and $0$ if feature $j$ is off. The equation we try to fit is given below:
%
\begin{equation}
	\tilde{\sigma} = a + F b + \epsilon
	\label{eqn:lin-regression}
\end{equation}
%
Linear regression finds the constant bias $a$ and weight vector $b$ that minimises the mean squared prediction error $\epsilon^T \epsilon / n$. The theory for linear regression is well known so omitted here (see \cite{yale-linear-regression} for a reference). We restrict $F$ to only include gender and language features and plot the corresponding element of $b$ for each feature in order of decreasing magnitude (figure \ref{fig:abp-output-lin-regression}).
%
\begin{figure}[H]
	\includegraphics[width=\linewidth]{ego-0-abp-lin-regression.png}
	\caption{Egonet-0 linear regression of $\tilde{\sigma}_v$ on select features ($R^2 = 0.113$)}
	\label{fig:abp-output-lin-regression}
\end{figure}

@@

This analysis is of course rather crude as we have failed to take into account multi-collinearity or even determined if a linear regression model is appropriate; indeed, an $R^2$ value of just 0.113 would suggest that it is not. Nevertheless, this serves as the starting point for subsequent more rigorous analysis. Furthermore, we have not yet proven the conjecture that the regression coefficients are valid measures of causal impact on graphical structure.

\pagebreak
\section{Future Direction}

The immediate next steps for the project are to refine the classification approach outlined in section \ref{sect:detection-to-inference}. Firstly, it should be tied back to the hypothesis testing framework to prove the results are be meaningful. Secondly, the exact form of the classifier can be explored; a linear regression model is a good starting point but we can expand to a simple softmax neural network as that would allow us to generalise to more than just two partitions.

Indeed, ABP is severely limited by the fact that it only outputs 2 sets. Though weak recovery is solved, the partition is often hard to interpret as multiple clusters can get subsumed into one. We instead would prefer to investigate algorithms that can identify arbitrarily many clusters. Force-directed approaches are the standard in industry for their speed of execution. One of the most common is the Fruchterman-Reingold algorithm which treats vertices as repelling objects and edges as restoring forces. It can then perform iterative updates on vertex positions until convergence or termination. This algorithm is what is used by the NetworkX package \cite{networkx} to compute graph layouts as seen in figures \ref{fig:ego-0-by-gender} and \ref{fig:abp-output-norm}. Song and Bressan \cite{force-directed} have had great success using the Fruchterman-Reingold algorithm followed by k-means clustering for community detection.

Lastly, the analysis must of course be expanded to more datasets to benchmark efficacy against existing methods and ensure the approach is generally applicable. Work has already begun analysing the influence of gender on academic collaboration graphs provided by Aminer \cite{aminer}. Furthermore, the seminal dataset of political blogs from the 2004 US election \cite{polblogs} which started much of the analysis of the SBM is in the works.

@@

We began by fully fleshing out the theoretical foundation of the Stochastic Block Model and hypothesis tests on connectivity parameters. Nevertheless, this approach is too narrow in scope to be a useful tool on its own. We therefore set about the problem from the opposite direction; rather than verify structure we are already given, we detect structure in the graph without using labels and seek to identify which labels best recreate the detected partition. Early results are promising but this approach needs to be given the same rigorous treatment as the hypothesis testing framework. That will be the emphasis of future work.
